---
title: "PT and MPT Implementation in Cpp test"
author: "Anh Phan"
date: "4/25/2024"
output: pdf_document
---

# Library needed

```{r}
library(parallel)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(mosaic)
library(Rcpp)
sourceCpp("~/PT_MPT/PT_MPT.cpp")
conflicted::conflicts_prefer(base::sum)
conflicted::conflicts_prefer(base::prod)
conflicted::conflicts_prefer(base::mean)
```


# Now, we can use this in an one sample analysis

We will use the mtcars data for this example, specifically the miles/gallon value of car

```{r}
data <- (mtcars$mpg)
# randomly chosen
# mu_0 <- 15
# sigma_0 <- 5
# how can it improved a informed parans
mu_0 <- mean(data)
sigma_0 <- sd(data)
```

We first consider the normal model for this data, assuming the mean and standard deviation equal the usual mean and standard deviation:

```{r}
normal <- data.frame(x = rnorm(1000, mean = mu_0, sd = sigma_0)) %>%
  mutate(y = dnorm(x, mean = mu_0, sd = sigma_0))
p <- ggplot() +
  geom_histogram(data = data.frame(data), aes(x = data), bins = 30) +
  ylab("Count")
p + geom_line(data = normal, aes(x = x, y = y *50, color = "Normal")) +
  scale_y_continuous(sec.axis = sec_axis(~ ./50, name = "Density")) +
  scale_color_manual(name = "Legend", 
                      values = c("Normal" = "red"))
```

Now we can test it on different Polya Tree prior

## Test on 1 level Polya tree prior with c = 1 with data update

We first try to draw random theta and then draw pdf on that

```{r}
pt_1_post <- pt_update_cpp(data, mu_0, sigma_0, J = 1)
pt_1 <- data.frame(x = rnorm(1000, mean = mu_0, sd = sigma_0))
# Now we draw 10 different distribution from PT prior
for (i in 1:10) {
  # get the column name to add into the data
  curr_col_name <- paste("y_", as.character(i), sep = "") 
  # now draw the random theta
  curr_theta_list <- draw_theta_cpp(TRUE, pt_1_post, 1, 1)
  # Now we fill the column
  curr_col <- unlist(lapply(pt_1$x, function(x) {return(pdf_pt_prior_cpp(x, mu_0, sigma_0, curr_theta_list))}))
  pt_1[,curr_col_name] <- curr_col
}
p <- ggplot() +
  geom_histogram(data = data.frame(data), aes(x = data), bins = 30) +
  ylab("Count")
for (i in 1:10) {
  # get the column name
  p <- p + geom_line(data = pt_1, aes_string(x = "x", y = paste("y_", as.character(i), "*50",sep = "")), linetype = "dashed")
}
p + scale_y_continuous(sec.axis = sec_axis(~ ./50, name = "Density"))
```

Now we can consider the mixture case:

```{r}
pt_1 <- data.frame(x = rnorm(1000, mean = mu_0, sd = sigma_0), 
                   y = numeric(1000))
for (i in 1:1000) {
  if (i %% 10 == 0) {
    # print(i)
  }
  pt_1[i, 2] <- pdf_pt_prior_mixture_c_cpp(pt_1[i, 1], mu_0, sigma_0, 1, 1, 10000, TRUE, pt_1_post)
}
p <- ggplot() +
  geom_histogram(data = data.frame(data), aes(x = data), bins = 30) +
  ylab("Count")
p + geom_line(data = normal, aes(x = x, y = y *50, color = "Normal")) +
  geom_line(data = pt_1, aes(x = x, y = y *50, color = "PT 1 level"))  + 
  scale_y_continuous(sec.axis = sec_axis(~ ./50, name = "Density")) +
  scale_color_manual(name = "Legend", 
                      values = c("Normal" = "red", 
                                 "PT 1 level" = "blue")) + 
  geom_vline(xintercept = mosaic::fav_stats(data)$Q1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = mosaic::fav_stats(data)$median, linetype = "dashed", color = "black") +
  geom_vline(xintercept = mosaic::fav_stats(data)$Q3, linetype = "dashed", color = "black")
  
```


## Now we will try a 2 level Polya tree prior instead of 1 level

```{r}
pt_2_post <- pt_update_cpp(data, mu_0, sigma_0, J = 2)
pt_2 <- data.frame(x = rnorm(1000, mean = mu_0, sd = sigma_0))
# Now we draw 10 different distribution from PT prior
for (i in 1:10) {
  # get the column name to add into the data
  curr_col_name <- paste("y_", as.character(i), sep = "") 
  # now draw the random theta
  curr_theta_list <- draw_theta_cpp(TRUE, pt_2_post, 1, 2)
  # Now we fill the column
  curr_col <- unlist(lapply(pt_2$x, function(x) {return(pdf_pt_prior_cpp(x, mu_0, sigma_0, curr_theta_list))}))
  pt_2[,curr_col_name] <- curr_col
}
p <- ggplot() +
  geom_histogram(data = data.frame(data), aes(x = data), bins = 30) +
  ylab("Count")
for (i in 1:10) {
  # get the column name
  p <- p + geom_line(data = pt_2, aes_string(x = "x", y = paste("y_", as.character(i), "*50",sep = "")), linetype = "dashed")
}
p + scale_y_continuous(sec.axis = sec_axis(~ ./50, name = "Density"))
```

Now we can consider the mixture case:

```{r}
pt_2 <- data.frame(x = rnorm(1000, mean = mu_0, sd = sigma_0), 
                   y = numeric(1000))
for (i in 1:1000) {
  if (i %% 10 == 0) {
    # print(i)
  }
  pt_2[i, 2] <- pdf_pt_prior_mixture_c_cpp(pt_2[i, 1], mu_0, sigma_0, 1, 2, 10000, TRUE, pt_2_post)
}
p <- ggplot() +
  geom_histogram(data = data.frame(data), aes(x = data), bins = 30) +
  ylab("Count")
p + geom_line(data = normal, aes(x = x, y = y *50, color = "Normal")) +
  geom_line(data = pt_2, aes(x = x, y = y *50, color = "PT 2 level"))  + 
  scale_y_continuous(sec.axis = sec_axis(~ ./50, name = "Density")) +
  scale_color_manual(name = "Legend", 
                      values = c("Normal" = "red", 
                                 "PT 2 level" = "blue")) + 
  geom_vline(xintercept = mosaic::fav_stats(data)$Q1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = mosaic::fav_stats(data)$median, linetype = "dashed", color = "black") +
  geom_vline(xintercept = mosaic::fav_stats(data)$Q3, linetype = "dashed", color = "black")
```


It seems polya tree would fit the data better based on the fact that it is an extension of its original distribution. Now, we can try to impose a distribution on mu and sigma, and then we take the mean to get the mixture of polya tree.

## Try with a Mixture of Polya Tree prior with 1 level

We will assume here that:

$$\mu \sim N(\mu_0, (\frac{\sigma_0}{10})^2)$$
and 
$$\sigma \sim N(\sigma_0, (\frac{\sigma_0}{10})^2)$$

```{r}
# We started with 1 level first
mpt_1 <- data.frame(x = rnorm(1000, mean = mu_0, sd = sigma_0))
# Our prior for mu and sigma
d_mu_test <- function(x) {
  return(dnorm(x, mu_0, sigma_0 / 10))
}
r_mu_test <- function() {
  return(rnorm(1, mu_0, sigma_0 / 10))
}
d_sigma_test <- function(x) {
  return(dnorm(x, sigma_0, sigma_0 / 10))
}
r_sigma_test <- function() {
  return(rnorm(1, sigma_0, sigma_0 / 10))
}
# Draw 10 random theta
for (i in 1:10) {
  print(i)
  # get the column name to add into the data
  curr_col_name <- paste("y_", as.character(i), sep = "") 
  # now draw the random theta
  curr_theta_list <- draw_theta_cpp(TRUE, pt_1_post, 1, 1)
  # Now we fill the column
  curr_col <- unlist(mclapply(mpt_1$x, 
                              function(x) {return(pdf_mpt_prior_cpp(x, 
                                                                mu_0, d_mu_test, r_mu_test, 
                                                                sigma_0, d_sigma_test, r_sigma_test,
                                                                curr_theta_list,
                                                                10000,
                                                                TRUE, data,
                                                                1, 1))},
                              mc.cores = 4))
  mpt_1[,curr_col_name] <- curr_col
}
p <- ggplot() +
  geom_histogram(data = data.frame(data), aes(x = data), bins = 30) +
  ylab("Count")
for (i in 1:10) {
  # get the column name
  p <- p + geom_line(data = mpt_1, aes_string(x = "x", y = paste("y_", as.character(i), "*50",sep = "")), linetype = "dashed")
}
p + scale_y_continuous(sec.axis = sec_axis(~ ./50, name = "Density"))
```

Seems like it is pretty smooth, how about a 2 level this time:

## Try with a Mixture of Polya Tree prior with 2 level

```{r}
mpt_2 <- data.frame(x = rnorm(1000, mean = mu_0, sd = sigma_0))
# Draw 10 random theta
for (i in 1:10) {
  print(i)
  # get the column name to add into the data
  curr_col_name <- paste("y_", as.character(i), sep = "") 
  # now draw the random theta
  curr_theta_list <- draw_theta(TRUE, pt_2_post, 1, 2)
  # Now we fill the column
  curr_col <- unlist(mclapply(mpt_2$x, 
                              function(x) {return(pdf_mpt_prior_cpp(x, 
                                                                mu_0, d_mu_test, r_mu_test, 
                                                                sigma_0, d_sigma_test, r_sigma_test,
                                                                curr_theta_list,
                                                                10000,
                                                                TRUE, data,
                                                                1, 2))},
                              mc.cores = 4))
  mpt_2[,curr_col_name] <- curr_col
}
p <- ggplot() +
  geom_histogram(data = data.frame(data), aes(x = data), bins = 30) +
  ylab("Count")
for (i in 1:10) {
  # get the column name
  p <- p + geom_line(data = mpt_2, aes_string(x = "x", y = paste("y_", as.character(i), "*50",sep = "")), linetype = "dashed")
}
p + scale_y_continuous(sec.axis = sec_axis(~ ./50, name = "Density"))
```

